Unrestricted Termination and Non-Termination Arguments for Bit-Vector Programs\thanks{Supported by UK EPSRC EP/J012564/1 and ERC project 280053.}

### 2015-01-16 papers/termination/termination-esop/paper.tex
added amir's commens

\begin{figure*}
\centering
 \begin{tabular}{|l||c|c|c|c|c|c|c|c|}
 \hline
                   & \multicolumn{8}{c|}{Program} \\
  Ranking argument & \multicolumn{2}{c|}{Rationals/Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
                   & L & NL & L & NL & L & NL & L & NL \\
  \hline
  \hline
  Linear lexicographic        & \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} & -- & \cite{DBLP:conf/tacas/LeikeH14} & -- &\checkmark&\checkmark&\checkmark&\checkmark\\
  Linear non-lexicographic    & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & \cite{DBLP:conf/tacas/LeikeH14} & -- & \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
  Nonlinear lexicographic     & -- & -- & -- & -- &\checkmark&\checkmark&\checkmark&\checkmark\\
  Nonlinear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} & \cite{DBLP:conf/vmcai/BradleyMS05} & -- & -- &\checkmark&\checkmark&\checkmark&\checkmark\\
  \hline
 \end{tabular}\\
 Legend: \checkmark = we can handle; --~=~no available works; L = linear; NL = nonlinear

@@

\item In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, program (i)
admits a \emph{multiphase} ranking function, computed from a multiphase
ranking template.  Multiphase ranking templates are targeted at programs
that go through a finite number of phases in their execution.  Each phase is
ranked with an affine-linear function and the phase is considered to be
completed once this function becomes non-positive.

In our setting this type of programs does not need special treatment, as we
can find a nonlinear lexicographic ranking function $R(x, y, z) = (x < y,
z)$.\footnote{This termination argument is somewhat subtle.  The Boolean
values $\mathit{false}$ and $\mathit{true}$ are interpreted as 0 and 1,
respectively.  The Boolean $x < y$ thus eventually decreases, that is to say
once a state with $x \geq y$ is reached, $x$ never again becomes greater
than $y$.  This means that as soon as the ``else'' branch of the if
statement is taken, it will continue to be taken in each subsequent
iteration of the loop.  Meanwhile, if $x < y$ has not decreased (i.e., we
have stayed in the same branch of the ``if''), then $z$ does decrease. 
Since a Boolean only has two possible values, it cannot decrease
indefinitely.  Since $z > 0$ is a conjunct of the loop guard, $z$ cannot
decrease indefinitely, and so $R$ proves that the loop is well founded.}

>>>>

\begin{figure*}
\centering
 \begin{tabular}{|l||c|c|c|c|c|c|c|c|}
 \hline
                   & \multicolumn{8}{c|}{Program} \\
  Ranking argument & \multicolumn{2}{c|}{Rationals/Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
                   & L & NL & L & NL & L & NL & L & NL \\
  \hline
  \hline
  Linear lexicographic        & \cite{DBLP:journals/jacm/Ben-AmramG14,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} & -- & \cite{DBLP:conf/tacas/LeikeH14} & -- &\checkmark&\checkmark&\checkmark&\checkmark\\
  Linear non-lexicographic    & \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & \cite{DBLP:conf/tacas/LeikeH14} & -- & \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
  Nonlinear lexicographic     & -- & -- & -- & -- &\checkmark&\checkmark&\checkmark&\checkmark\\
  Nonlinear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} & \cite{DBLP:conf/vmcai/BradleyMS05} & -- & -- &\checkmark&\checkmark&\checkmark&\checkmark\\
  \hline
 \end{tabular}\\
 Legend: \checkmark = we can handle; --~=~no available works; L = linear; NL = nonlinear

@@

\item %% In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, program (i)
%% admits a \emph{multiphase} ranking function, computed from a multiphase
%% ranking template.  Multiphase ranking templates are targeted at programs
%% that go through a finite number of phases in their execution.  Each phase is
%% ranked with an affine-linear function and the phase is considered to be
%% completed once this function becomes non-positive.

%% In our setting this type of programs does not need special treatment, as
For program (i) we find a nonlinear lexicographic ranking function $R(x, y, z) = (x < y,
z)$.\footnote{This termination argument is somewhat subtle.  The Boolean
values $\mathit{false}$ and $\mathit{true}$ are interpreted as 0 and 1,
respectively.  The Boolean $x < y$ thus eventually decreases, that is to say
once a state with $x \geq y$ is reached, $x$ never again becomes greater
than $y$.  This means that as soon as the ``else'' branch of the if
statement is taken, it will continue to be taken in each subsequent
iteration of the loop.  Meanwhile, if $x < y$ has not decreased (i.e., we
have stayed in the same branch of the ``if''), then $z$ does decrease. 
Since a Boolean only has two possible values, it cannot decrease
indefinitely.  Since $z > 0$ is a conjunct of the loop guard, $z$ cannot
decrease indefinitely, and so $R$ proves that the loop is well founded.}
 We are not aware of any linear ranking function for this program.
